---
title: "MBA-6361[C]"
author: "Prince Batra"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: simplex
    code_folding: hide
    toc: true
    toc_float: true
---

```{r Loading Libraries, echo=TRUE, warning=FALSE, message=FALSE, error=TRUE}

# Load libraries


library(kableExtra)
library(dplyr)
library(tidyr)
library(broom)
library(lubridate)
library(knitr)
library(readr)
library(ggplot2)
library(stringr)
library(magrittr)
library(markdown)
library(purrr)
library(rmarkdown)
library(tibble)
library(readxl)
library(janitor)
library(ggrepel)
library(skimr)
library(scales)
library(patchwork)
library(tictoc)

```

```{r setup 1, include=FALSE, echo = FALSE, warning=FALSE, error=TRUE}
knitr::opts_chunk$set(echo = TRUE, fig.height = 5, fig.width = 5)
```

# 1 - Problem Statement

Going even deeper than before to find insights from the elections held on Oct 24, 2018 with a wider dataset and presenting results found during the process

# 2 - Is Everyone Really Voting?

Please see the code used for this section as required

```{r Global Variable, echo=TRUE, warning=FALSE, message=FALSE}

Score_Card <- data.frame()

Summary_Election_Data <- read_excel("statementofvotescastoctober242018.xls", sheet = 1, skip = 2) %>%
  clean_names() %>%
  filter(!is.na(registered_voters)) %>%
  select_if(~ !any(is.na(.))) %>%
  filter(str_detect(precinct, "^0|^1|^2"))
```

```{r Defining Function, Cleaning Data & Performing Calculations, echo=TRUE, message=FALSE, error=TRUE }


cal_election_data <- function(filename, Ward_No, tibble = TRUE)
  
{

  
  Election_Data <-
    read_excel("statementofvotescastoctober242018.xls",
               sheet = Ward_No ,
               skip = 2) %>%
    clean_names() %>%
    filter(!is.na(times_cast)) %>%
    select_if(~ !any(is.na(.))) %>%
    select (-precinct_1:-overvotes) %>%
    rename("precinct" = "precinct_7") %>%
    filter(str_detect(precinct, "^0|^1|^2")) 
    

  
  ELection_Data_Consolidated <-
    merge(Election_Data, Summary_Election_Data[c("precinct", "registered_voters")], by =
            "precinct") %>% adorn_totals("row")
  
  

  
  for (i in 2:ncol(ELection_Data_Consolidated) - 1)
  {
    if (ELection_Data_Consolidated[nrow(ELection_Data_Consolidated), i] == max(ELection_Data_Consolidated[nrow(ELection_Data_Consolidated), 2:(ncol(ELection_Data_Consolidated) -
                                                                                                                                               2)]))
    {
      Election_Data_Mini <- data.frame(ELection_Data_Consolidated[i])
      colnames(Election_Data_Mini) <-
        colnames(ELection_Data_Consolidated[i]) 
    }
  }
  
 
  
  Final_Election_Data <-
    data.frame(
      Ward_Number = Ward_No-2,
      Ward_Winner = colnames(Election_Data_Mini),
      Winner_Votes = as.integer(Election_Data_Mini[nrow(Election_Data_Mini), ncol(Election_Data_Mini)]),
      Voters_Voted = as.integer(ELection_Data_Consolidated[nrow(ELection_Data_Consolidated), ncol(ELection_Data_Consolidated) -
                                                             1]),
      Registered_Voters = as.integer(ELection_Data_Consolidated[nrow(ELection_Data_Consolidated), ncol(ELection_Data_Consolidated)]),
      Voters_Voted_Winner_Fraction = round(
        as.numeric(Election_Data_Mini[nrow(Election_Data_Mini), ncol(Election_Data_Mini)] /
                     ELection_Data_Consolidated[nrow(ELection_Data_Consolidated), ncol(ELection_Data_Consolidated) -
                                                  1]),
        digit = 3
      ),
      Voters_Voted_Ward_Fraction = round(
        as.numeric(ELection_Data_Consolidated[nrow(ELection_Data_Consolidated), ncol(ELection_Data_Consolidated) -
                                                1] / ELection_Data_Consolidated[nrow(ELection_Data_Consolidated), ncol(ELection_Data_Consolidated)]),
        digit = 3
      ),
      stringsAsFactors = FALSE
    )
  
  return(Final_Election_Data)
  
}
```

```{r Generating Election Score Card, echo=TRUE, message=FALSE, warning=FALSE, error=TRUE }


for (X in 3:25)
{
  Score_Card <-
    rbind(Score_Card,
          cal_election_data("statementofvotescastoctober242018.xls", X))
}

```

```{r Final Analysis, echo=TRUE, error=TRUE, message=FALSE, warning=FALSE}


Census_Data <-
  read_csv("census_2016.csv", col_names = TRUE) %>%
  clean_names()

Ward_Info <-
  read_csv("ward_info_2011.csv", col_names = TRUE) %>%
  clean_names()



Long_Score <-
  merge(Census_Data, Ward_Info[, c("number", "area")], by = "number")
Merged_Score <-
  merge (Score_Card, Long_Score, by.x = "Ward_Number", by.y = "number")



Merged_Score$Avg_People_Per_House <-
  round(Merged_Score$population / Merged_Score$households, digits = 2)
Merged_Score$Population_Density <-
  round(Merged_Score$population / Merged_Score$area, digit = 2)
Merged_Score$Ratio_Fraction_Winner_Ward_Votes <-
  round(
    Merged_Score$Voters_Voted_Winner_Fraction / Merged_Score$Voters_Voted_Ward_Fraction,
    digits = 2
  )



for (i in 1:nrow(Merged_Score))
{
  if (Merged_Score$Population_Density[i] > quantile(Merged_Score$Population_Density, probs = 0.66666))
  {
    Merged_Score$Ward_Type[i] <- "Urban"
  }
  else if (Merged_Score$Population_Density[i] <= quantile(Merged_Score$Population_Density, probs = 0.66666) &&
           Merged_Score$Population_Density[i] >= quantile(Merged_Score$Population_Density, probs = 0.33333))
  {
    Merged_Score$Ward_Type[i] <- "Suburban"
  }
  else
  {
    Merged_Score$Ward_Type[i] <- "Rural"
  }
}


```



```{r setup 2, include=FALSE, echo = FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.height = 7, fig.width = 7)
```

## 2.1 - Congratulations Winners But?

The following graph shows the winners from each Ward along with the fraction of Votes, out of Total Votes in that Ward, the winner received.

However, it doesn't really present a clear picture about the actual number of voters that participated in the election.

```{r graph 1, echo=TRUE, message=FALSE, warning=FALSE, error=TRUE }



Merged_Score %>%
  ggplot(aes(x=Ward_Winner,y=Voters_Voted_Winner_Fraction, fill=as.factor(ward)))+
  geom_bar(stat="identity",position="stack")+
  coord_flip()+
    labs(
    x = 'Ward_Winner',
    y = 'Fraction of Votes',
    fill = "Ward Name",
    title = 'Winners in Each Ward'
  )
```

## 2.2 - Actual Participation in Elections

Please see the data as shown below for reference

```{r Data Table, echo=TRUE, message=TRUE, warning=FALSE, error=TRUE }
kable(Merged_Score) %>%
  kable_styling("striped", full_width = F) %>%
  scroll_box(width = "700px", height = "400px")
```



Based upon the above data we plotted the below graph which details the ratio between fraction of votes winner received out of Total votes given in the ward, to, fraction of people that voted out of total number of people that were eligible to vote.


  ```{r Graph 2, echo=TRUE, message=TRUE, warning=FALSE, error=TRUE }

a<-coef(lm(Ratio_Fraction_Winner_Ward_Votes ~ Avg_People_Per_House, data = Merged_Score)) 


Merged_Score %>%
  ggplot(
    aes(
      x = Avg_People_Per_House,
      y = Ratio_Fraction_Winner_Ward_Votes,
      colour = Ward_Type
    )
  ) +
  geom_point(size=2) +
  labs(x = 'People per Household',
       y = 'Ratio between Fraction of Winner Votes and Fraction of Eligible Voters',
       colour = 'Ward Type') +
  scale_x_continuous(labels = comma) +
  scale_y_continuous(labels = comma) +
  geom_smooth(method = 'lm',se=FALSE) +
  geom_abline(
    slope = a[2],
    intercept = a[1],
    colour = 'red',
    size = 3,
    alpha = 0.1
  )

```

## 2.3 - Insights Incurred:

Performing this analysis gave many insights as follows:

* One trend we could see from this graph is more the number of people in a house, lesser the number of votes the winner gets. However it maybe due to less number of eligible voters in the house (for example kids)
* The second detail to otice is high performance from Rural areas which could be contributed to less population and hence overall high number of eligible voters
* The general trend whether we look at Rural, Suburban or Urban areas, is a decrease in participation in elections.
* From the Data Table, participation from eligible varied with the lowest level of participation from Rideau-Vanier and the highest particiaption from Capital where about 40% of the eligible voters gave votes in the election.

## 2.4 - Final Comments:

Steps should be taken to motivate eligible voters to vote. As if all eligible voters had votes, the result may have been different in some of the Wards where the winning margin was quite close.

# 3 - Next Steps

* An interesting analysis would be to calculate the rate of population growth over years in Ottawa and see how that relates with rate of voters voting. Is it decreasing or is it increasing?
* Another analysis that I would like to do is to check increase in number of households and average people per household and see how it relates with people voting and rate of population growth

# 4 - References

* For Section 2 Titled "Analysis of Data", Ideas taken from https://statisticsglobe.com/r-max-min-function/
* For Section 2 Titled "Analysis of Data", Taken tips on how to better used variables in a function and how to return them     https://community.rstudio.com/t/access-local-variable-in-global-scope-outside-function/26616/3
* For Section 2 Titled "Analysis of Data", Ideas taken to develop a way to read all sheets effectively.
  https://stackoverflow.com/questions/12945687/read-all-worksheets-in-an-excel-workbook-into-an-r-list-with-data-frames


# 5 - Appendix: sessionInfo

```{r}
sessionInfo()
```
